---
title: "bacteria response tutorial"
author: "Putt, Andrew"
date: "5/27/2022"
this tutorial uses tools provided through phyloseq, microbiota processes, and others and are not owned, maintained, or created by the author
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.14")
library(BiocManager)

# phyloseq installer
source("https://raw.githubusercontent.com/joey711/phyloseq/master/inst/scripts/installer.R",
       local = TRUE)
# install most recent version of phyloseq from github 
install_phyloseq(branch = "github")
library(phyloseq)

list.of.packages <- c("plyr","dplyr","tidyr","knitr","corrplot","ggcorrplot","vegan",
                      "tidyverse","tidytree","RColorBrewer","metacoder","ggpubr","ggplot2",
                      "ape","readr","foreach","doParallel","scales","grid","reshape2",
                      "multcompView", "viridis","VennDiagram","UpSetR","RColorBrewer",
                      "phyloseq","pheatmap", "base","plotly","vegan","ggpubr","ggtree", 
                      "ggfun","ggjoy","data.table","ggstance")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages) # install
lapply(list.of.packages, require, character.only = TRUE) #  load

# install bioconductor packages
Bioc.packages <- c("microbial","microbiome","MicrobiotaProcess","DESeq2","dada2",
                   "DECIPHER","edgeR", "ANCOMBC","lattice")
new.Bioc.packages <- Bioc.packages[!(Bioc.packages %in% installed.packages()[,"Package"])]
if(length(new.Bioc.packages)) BiocManager::install(new.Bioc.packages) #install
lapply(Bioc.packages, require, character.only = TRUE) # load
```

#load data from dropbox directory
```{r}
#load the data (requires a stable internet connection)
otu_2017 <-as.data.frame(read.csv("https://www.dropbox.com/s/qmcar6i70s95zub/EVO_community_2017_all_final.csv?dl=1", sep = ",", row.names = 1, stringsAsFactors = FALSE ))
tree_2017 <- read_tree("https://www.dropbox.com/s/82oabfa65inq2ud/EVO_tree_2017_all_final.nwk?dl=1")
taxa_2017 <-as.data.frame(read.csv("https://www.dropbox.com/s/q1erh73ml6e1ovi/EVO_clasf_2017_all_final.csv?dl=1", sep = ",", row.names = 1, stringsAsFactors = FALSE ))
# sample data
meta_complete <- read.csv("https://www.dropbox.com/s/tzce9ie41bj3pr0/Unified_EVO_metadata_sheet.csv?dl=1", row.names = 1,stringsAsFactors = FALSE)

```

#clean up the data
```{r}
#reassign sample data column names
colnames(meta_complete) <- c('WellID','Date','DaysPostinjection','FilterSize','EVOYear','well_id','CellCount',
                            'DissolvedOxygen _mgL','pH','Temperature C','Redox_mV','SpecificConductivity.uScm',
                            'watertable_famsl','flow_cms','Na_mgL','Mg_mgL', 'Al_mgL', 'K_mgL', 
                            'Ca42_mgL','Ca43_mgL', 'Ca44_mgL', 'Fe54_mgL', 'Mn_mgL', 'Fe57_mgL',
                            'U238_mgL', 'NO2_mgL', 'NO3_mgL', 'SO4_mgL', 'Lactate_uM', 'Acetate_uM', 
                            'Propionate_uM','Formate_uM', 'Butyrate_uM', 'Pyruvate_uM', 'Lithium_mgL', 
                            'NH4_mgL','well_type','Oxygen_content','nitrate_reduction','Days','sample_label')
glimpse(meta_complete)

# coerce taxonomy data into a phyloseq format
taxamat <- dplyr::select(taxa_2017, -ends_with(c(".Confidence"))) # remove confidence scores
names(taxamat)[names(taxamat) == "Kingdom"] <- "Domain" # phyloseq uses 'Domain' instead of 'Kingdom'
#remove missing
taxamat <- as.matrix(taxamat[!is.na(taxamat$Phylum),])
```

#make phyloseq object
```{r}
#build a phyloseq object
phylo2017 <- phyloseq(otu_table(otu_2017, taxa_are_rows = TRUE), phyloseq::tax_table(taxamat), 
                      sample_data(meta_complete), tree_2017)
# rarefy the 2017 phyloseq object
set.seed(0451)
phylo2017_rarefied<-rarefy_even_depth(phylo2017,sample.size = min(sample_sums(phylo2017)))
```

#restrict phyla to most abundant
```{r}
# phylum-specific changes
phylo2017_rarefied
#select phyla with >500 counts
prune_phylo2017 <-phyloseq::prune_taxa(taxa_sums(phylo2017_rarefied) > 500, phylo2017_rarefied)
#identify table with phylum-level IDs
table(phyloseq::tax_table(prune_phylo2017)[, "Phylum"])
# remove low-count OTUs from phylum subset
prune_phylo2017 <-phyloseq::subset_samples(sample_sums(prune_phylo2017) > 50, prune_phylo2017)
ps_phylum <- phyloseq::tax_glom(prune_phylo2017, "Phylum")
phyloseq::taxa_names(ps_phylum) <- phyloseq::tax_table(ps_phylum)[, "Phylum"]
#melt the phyloseq metadata using the subset
melt_pr_phylo2017_rarefied <- phyloseq::psmelt(prune_phylo2017)
#transform phylum-level phyloseq object
ps_phylum.df <- as.data.frame(t(otu_table(ps_phylum)))
colnames(ps_phylum.df)
#extract sample data related to the phyloseq
ps_samdat.df <- as.data.frame(sample_data(phylo2017_rarefied))
#combine sample data with pruned phyloseq data
ps_phylum_merge <- merge(x=ps_phylum.df, y=ps_samdat.df,
                         by = 'row.names', all = TRUE)
#phylum changes
ps_melt_df <- phyloseq::psmelt(ps_phylum)
ps_melt_df <- as.data.frame(ps_melt_df)
```

#two examples of comparison box plots with wilcoxon test p-values
```{r}
#example 1: comparison of control and monitoring groups
phyloseq::psmelt(ps_phylum) %>%
  ggplot(data = ., aes(x = well_type, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  theme_bw()+
  geom_jitter(aes(color = OTU), height = 0, width = .2) +
  labs(x = "", y = "Number of ASVs\n", color = "Phyla") +
  facet_wrap(~ OTU, scales = "free") +
  ggtitle("Changes in taxonomic abundance with well type") +
  theme(axis.title = element_text(size = 12),
        legend.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        strip.text.x = element_text(size = 12),
        axis.text = element_text(size = 12)) + stat_compare_means()
ggsave("~/phylum_welltype_boxplot_facet.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/phylum_welltype_boxplot_facet.png")
```

#example with nitrate reduction
```{r}
#example 2: comparison of abundant taxa with changes in nitrate levels
phyloseq::psmelt(ps_phylum) %>%
  ggplot(data = ., aes(x = nitrate_reduction, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  theme_bw()+
  geom_jitter(aes(color = OTU), height = 0, width = .2) +
  labs(x = "", y = "Number of ASVs\n", color = "Phyla") +
  facet_wrap(~ OTU, scales = "free") +
  ggtitle("Changes in taxonomic abundance with nitrate content") +
  theme(axis.title = element_text(size = 12),
        legend.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        strip.text.x = element_text(size = 12),
        axis.text = element_text(size = 12)) +stat_compare_means()
ggsave("~/phylum_nitrate_boxplot_facet.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/phylum_nitrate_boxplot_facet.png")
```

#changes in abundance with nitrate concentration and phyla
```{r}
phyloseq::psmelt(ps_phylum) %>%
  ggplot(data = ., aes(x = NO3_mg.L, y = Abundance)) +
  geom_smooth(aes(color=OTU))+
  geom_jitter(aes(shape =FilterSize,color=OTU), height = 0, width = .2) +
  theme_bw()+
  geom_jitter(aes(color = OTU), height = 0, width = .2) +
  labs(x = "Nitrate (mg/L)", y = "Number of ASVs\n", color = "Phyla") +
  facet_wrap(~ OTU, scales = "free") +
  ggtitle("Changes in taxonomic abundance with nitrate content") +
  theme(axis.title = element_text(size = 12),
        legend.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        strip.text.x = element_text(size = 12),
        axis.text = element_text(size = 12))
ggsave("~/nitrate_lineplot_phylum_facet.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_lineplot_phylum_facet.png")
```

#lineplot with facet by filter size
```{r}
ggplot(data = ps_melt_df, aes(x = NO3_mg.L, y = Abundance)) +
  theme_bw()+
  geom_smooth(aes(color=OTU))+
  geom_jitter(aes(shape =FilterSize,color=OTU), height = 0, width = .2) +
  labs(x = "Nitrate (mg/L)", y = "Number of ASVs\n", color = "Phyla") +
  facet_grid(FilterSize~ OTU, scales = "free") +
  ggtitle("Changes in taxonomic abundance with Nitrate Concentration") +
  theme(axis.title = element_text(size = 12),
        legend.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        strip.text.x = element_text(size = 12),
        axis.text = element_text(size = 12))
ggsave("~/nitrate_lineplot_phylum_filter_facet.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_lineplot_phylum_filter_facet.png")
```

#ensure that the phyloseq object is assigned properly
```{r}
phylo2017_rarefied_prune <- prune_taxa(taxa_sums(phylo2017_rarefied)>500, phylo2017_rarefied)
phylo2017_rarefied_prune <- tax_glom(phylo2017_rarefied_prune, "Family")
phylo2017_rarefied_prune
melt_simple <- psmelt(phylo2017_rarefied_prune) %>% 
  filter(Abundance < 10) %>% 
  select(OTU, val=Abundance)
#view the 'rarefied' melted dataframe
head(melt_simple)
```

#make a simple phylum-level tree
```{r}
p <- ggtree(phylo2017_rarefied_prune, layout = "rectangular") + 
  geom_tippoint(aes(color = Phylum),
    size = 2.5, alpha = 0.5) 
p
```

#merge tree with abundance density plots
```{r}
p2 <- facet_plot(p, panel="Abundance", data=melt_simple, 
           geom_density_ridges, mapping = aes(x=val,group=label, 
                                   fill=Phylum),  
           color='grey80', lwd=.3)
p2
ggsave("~/abundance_tree.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/abundance_tree.png")
```

#subsetting a phyloseq object
```{r}
# keep only proteobacteria
p_2017_proteo <- subset_taxa(phylo2017_rarefied_prune, Phylum=="Proteobacteria")
# filter proteobacteria by those found in the well MLSB3 with abundant taxa identified from each day from day -6 to day 200
sample_data(p_2017_proteo)$MLSB3 <- get_variable(p_2017_proteo, "Days") %in% 
  c(-6:200) 
#merging samples to a single dataframe based on well ID which in this case is 'MLSB3'
mergedMLBS <- merge_samples(p_2017_proteo, "WellID")

colnames(meta_2017)
```

#merge a ggplot object (geom_point()) into a ggplot
```{r}
ggtree(mergedMLBS, layout = "rectangular") + 
  geom_nodelab(aes(label=label), hjust=-.05, size=3.5) +
  
  geom_point(aes(x=x+hjust, color=NO3_mg.L, 
                 size=Abundance), alpha = 0.6, na.rm=TRUE) +
  geom_tiplab(aes(label=Family), hjust=-.5, size = 8)+   
  #scale_color_hue(l = 45) +
  scale_size_continuous(trans=log_trans(5)) + ggtitle("Abundant Proteobacteria Families in Well MLSB3") +
  theme(legend.position="right", title = element_text(face = "bold")) + hexpand(.4)
ggsave("~/nitrate_MLBS_proteo_tree.png", 
       bg = "transparent", width = 40, height = 25, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_MLBS_proteo_tree.png")
```

#Linear Discriminate Analysis (LDA)
```{r}
# ggdiffclade 
set.seed(100)
Nitrate_diff <- diff_analysis(obj = phylo2017_rarefied, classgroup = "nitrate_reduction",
                       mlfun = "lda", filtermod = "pvalue", firstcomfun = "kruskal.test",
                       firstalpha = 0.05, strictmod = TRUE, secondcomfun = "wilcox.test",
                       subclmin = 3, subclwilc = TRUE, secondalpha = 0.01, lda=3)
# print out results of the diff_analysis                     
nitrate_diff_result <- as.data.frame(Nitrate_diff@result)
write.csv(nitrate_diff_result, "~/nitrate_diff_result.csv")
```

#plot differential analysis results with abundance of taxa and linear discriminate analysis model 'effect size'
```{r}
# diff_clade plot
diffclade_N <- ggdiffclade(obj=Nitrate_diff, alpha=0.3, linewd=0.15, skpointsize=0.6,
                           layout="radial", taxlevel=3, removeUnknown=TRUE, reduce=TRUE # This argument is to remove the branch of unknown taxonomy.
) +
  scale_fill_manual(
    values=c("#00AED7", "#FD9347")
  ) +
  guides(color = guide_legend(
    keywidth = 0.1, keyheight = 0.6, order = 3, ncol=1)) +
  theme(panel.background=element_rect(fill=NA),legend.position="right", 
        plot.margin=margin(0,0,0,0), legend.spacing.y=unit(0.02, "cm"), legend.title=element_text(size=7),
        legend.text=element_text(size=6), legend.box.spacing=unit(0.02,"cm"))
diffclade_N
ggsave("~/nitrate_diffclade_color_tree_plot.tiff", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_diffclade_color_tree_plot.tiff")
```

#linear range plot (using ggdiffbox)
```{r}
diffbox_N <- ggdiffbox(obj=Nitrate_diff, box_notch=FALSE, 
                     colorlist=c("steelblue", "goldenrod"), l_xlabtext="relative abundance") 
diffbox_N
ggsave("~/nitrate_reduction_LDA_boxplot.tiff", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_reduction_LDA_boxplot.tiff")
```

#plotting just the effect size
```{r}
es_nitrate <- ggeffectsize(obj=Nitrate_diff, 
                          lineheight=0.1,
                          linewidth=0.3) + 
  scale_color_manual(values=c("darkorange3", "slateblue")) 
es_nitrate
ggsave("~/nitrate_reduction_LDA_effectsize_boxplot.png", 
       bg = "transparent", width = 60, height = 35, units = "cm")
cat("The plot has been been saved to your directory:",getwd(),"/nitrate_reduction_LDA_effectsize_boxplot.png")
```

## To run with your own data:
# 1. load the required packages
# 2. input your own data 
# 3. make your own phyloseq object
# 4. change the data labels and titles to match your specific data use colnames(my metadata) to view the column names for making your own aesthetics 'aes' in the ggplots to test categories of interest for your analysis.

## contact Andrew (aputt@vols.utk.edu) with questions or through github https://github.com/putt-ad to resolve issues
